---
title: "Data explorer"
---

```{r load_data, include=FALSE, message=FALSE, warning=FALSE}

Sys.setenv(LANG = "en")
library(lubridate)
library(stringi)
library(httr)
library(jsonlite)

url_api <- "https://api.github.com/repos/jatorresmunguia/disease_outbreak_news/contents/Last%20update"
last_file <- fromJSON(content(GET(url_api), as = "text"))$name[grepl(fromJSON(content(GET(url_api), as = "text"))$name, pattern = paste0("^outbreaks"))]

rdata_file <- last_file[grepl(".csv$", last_file)]
file_name <- basename(rdata_file)
date_string <- sub(".*_(\\d{2})(\\d{2})(\\d{4}).*", "\\1-\\2-\\3", file_name)
date_obj <- dmy(date_string)

formatted_date <- format(date_obj, "%d/%m/%Y")  
formatted_month <- format(date_obj, "%B %d, %Y")     

last_version <- read.csv(paste0("https://raw.githubusercontent.com/jatorresmunguia/disease_outbreak_news/refs/heads/main/Last%20update", "/", rdata_file),
                         row.names = 1, header = TRUE)

num_countries <- nlevels(factor(last_version$Country))
num_diseases <- nlevels(factor(last_version$icd104n))

```

::: {.callout-important appearance="simple" collapse="true"}
### Methodological note
The unit of analysis in the database is an **outbreak**, defined as the occurrence of at least one case of a specific disease in a country -or territory- during a particular year. Therefore, a country -or territory- cannot have more than one outbreak of the same disease in the same year, although it may experience outbreaks of different diseases within the same year. A country can only have multiple outbreaks of the same disease if they occur in different years.
:::

::: {.callout-tip appearance="simple" collapse="true"}
### Data records
**Country**: Name of the country where the outbreak occurred; **Year**: Year of occurrence of the outbreak; **Type of disease**: Type of disease according to the ICD-10; **Disease**: Name of disease according to the ICD-10.
:::

::: {.callout-note appearance="simple" collapse="true"}
### Time & geographic data coverage
The last version of the dataset contains information on **`r nrow(last_version)` outbreaks**: <br>
**Temporal coverage**: 01/01/1996 - `r formatted_date` <br>
**Geographic coverage**: `r num_countries` <br>
**Number of diseases included**: `r num_diseases`
:::

::: {.callout-caution appearance="simple" collapse="true"}
### How to use the data explorer?
Use the **search** filters at the top of each column to narrow down specific countries, years, or diseases. The search is case-insensitive and does not use regular expressions.<br>
You can **sort** the columns by clicking on the headers (e.g., Country, Year). By default, the data is ordered alphabetically by Country and chronologically by Year, from the most recent to the oldest. Scroll through the table to view additional rows of the dataset.<br>
You can **export** the data in various formats (.csv, .xlsx, PDF) or copy selected information for further analysis using the available tools.
:::

```{r table_outbreaks, include=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
library(DT)
library(tidyverse) 

last_version |> 
  select(Country, Year, icd10n, icd104n) |>
  datatable(colnames = c("Country", "Year", "Type of disease", "Disease"),  
            rownames = FALSE, 
            filter = 'top',
            class = 'row-border hover stripe',
            extensions = 'Buttons',  
            options = list(
              dom = 'Bfrtip', 
              buttons = c('copy', 'csv', 'excel', 'pdf'),
              order = list(list(0, 'asc'), list(1, 'desc'))
              )
            )

```